---
title: "Darwin Core mapping"
subtitle: "For: The (...) occurrences of "
author:
- author_1
- author_2
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
#  pdf_document:
#    df_print: kable
#    number_sections: yes
#    toc: yes
#    toc_depth: 3
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c("tidyverse", "tidylog", "magrittr", "here", "janitor", "readxl", "digest", "rgbif")
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
library(sf)
```

# Read source data

Create a data frame `input_data` from the source data:

```{r}
input_data <- read.csv(file = here("data", "raw", "rato_data.csv"),
                       header = T,
                       sep = ",") 
```

Preview data:

```{r}
input_data %>% head(n = 5)
```


New version raw data as downloaded via `read_data_from_wfs.R`:

```{r}
input_data_new <- readr::read_csv(
  here::here("data", "raw", "rato_data_new.csv")
) 
```

New columns have been added:

```{r}
names(input_data_new)[!names(input_data_new) %in% names(input_data)]
```

# Process source data

## Tidy data

Clean data somewhat:

```{r}
input_data <- 
  input_data %>% 
    remove_empty("rows") %>% 
    clean_names()
```

`object_id` is the unique identifier for each occurrence. This will be the `occurrenceID`
`dossier_id` is the unique identifier for each sampling event. This will be the `eventID`

Check whether `object_id` is unique (value should be `0`)

```{r}
anyDuplicated(input_data$dossier_id)
```

Remove all rows for `Domein` = `Werken` 

```{r}
input_data %<>% filter(domein != "Werken")
```
## Add eventID and occurrenceID

```{r}
input_data %<>% mutate(event_id = dossier_id)
```
# Event core

## Pre-processing

Create a dataframe with unique events only:

```{r}
event <- input_data %>% distinct(event_id, .keep_all = TRUE)
```

## Term mapping

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records).

### type

```{r}
event %<>% mutate(dwc_type = "Event")
```

### language

```{r}
event %<>% mutate(dwc_language = "en")
```

### license

```{r}
event %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
event %<>% mutate(dwc_rightsHolder = "RATO")
```

### datasetID

```{r}
event %<>% mutate(dwc_datasetID = "")
```

### institutionCode

```{r}
event %<>% mutate(dwc_institutionCode = "RATO")
```

### datasetName

```{r}
event %<>% mutate(dwc_datasetName = "RATO - daily operations commissioned by the province East Flanders, Belgium")
```

### basisOfRecord

```{r}
event %<>% mutate(dwc_basisOfRecord = "humanObservation")
```

The following terms contain information about the sampling event:

### eventID

```{r}
event %<>% mutate(dwc_eventID = event_id)
```

### recordedBy

```{r}
event %<>% mutate(dwc_recordedBy = "RATO")
```
### organismQuantity, organismQuantityType and occurrenceStatus

Information is included in `waarnemingen`

**splitsen obv separator**

```{r}
event %>% 
  group_by(waarneming) %>% 
  summarise(records = n())
```
Split into different columns: `waarneming_type` and `waarneming_kwantiteit`:

```{r}
event <- 
  event %>% 
    separate(
      col = waarneming,
      into = c("waarneming_type", "waarneming_kwantiteit"),
      sep = "\\s=\\s",
      remove = FALSE) %>% 
    mutate(waarneming_kwantiteit = str_remove(
      string = waarneming_kwantiteit,
      pattern = ";"))
```

```{r}
event %<>% 
  mutate(organismQuantity = case_when(
    waarneming_type == "Vastgesteld (in m²)" |
    waarneming_type == "Vastgesteld (aantal)" ~ waarneming_kwantiteit,
    TRUE ~ ""
  )) 
```

```{r}
event <-
  event %>% 
    mutate(organismQuantityType = case_when(
      waarneming_type == "Vastgesteld (in m²)" ~ "square meter(s)",
      waarneming_type == "Vastgesteld (aantal)" ~ "individual(s)",
      TRUE ~ ""
    ))
``` 
### occurrenceStatus

TBD, use logical operators


### samplingProtocol

```{r}
event %>% 
  group_by(materiaal_vast) %>% 
  summarise(records = n()) %>% 
  arrange (records)
```

First split on ";" and separate on "="

```{r}
sampling_protocol <-
  event %>% 
    select(event_id, materiaal_vast) %>% 
    # separate on ";" and split in different rows:
    separate_rows(
      "materiaal_vast",
      sep = ";\\s") %>% 
  
    # separate on " = " and split in different columns:
    separate(
      col = materiaal_vast,
      sep = "\\s=\\s",
      into = c("materiaal", "kwantiteit"))  %>% 
      
    # remove "(aantal)" for easier mapping:
   mutate(materiaal = str_remove(materiaal,"\\s\\(aantal\\)")) %>% 
  
    # remove all rows for which `kwantiteit = NA
  filter(!is.na(kwantiteit))
```

Translate `materiaal` to English:

```{r}
sampling_protocol %>% 
  select(materiaal) %>% 
  group_by_all() %>% 
  summarise(records = n())
```
Translate to English:

```{r}
sampling_protocol %<>% 
  mutate(protocol = recode(materiaal,
    "Andere" = "other",
    "Conibearklem" = "conibear trap(s)",
    "Fuik" = "fike(s)",
    "Grondklem" = "ground trap(s)",
    "Klemvlot" = "",
    "Lokaasklem" = "bait trap",
    "Materiaal gesaboteerd" = "",
    "Slagnet" = "",
    "Vangkooi" = "cage",
    "Vangnet" = "net",
    "Wildcamera" = "camera trap"))
```
Add sampling effort:

```{r}
sampling_protocol %<>% 
  mutate(effort = case_when(
    kwantiteit == "0" ~ "",
    TRUE ~ paste(protocol, kwantiteit, sep = "=")))
```

```{r}
sampling_protocol <-
  sampling_protocol %>%
    group_by(event_id) %>%
    summarise(materials = paste(protocol, collapse = " | "),
              materials_effort = paste(effort, collapse = " | "))
```
Merge dataset:

```{r}
event %<>% left_join(
  y = sampling_protocol,
  by = "event_id")
```


### samplingEffort

### eventDate

```{r}
event %<>% mutate(dwc_eventDate = as.Date(aangemaakt_datum))
```

### countryCode

```{r}
event %<>% mutate(dwc_countryCode = "BE") 
```

### verbatimLatitude

```{r}
event %<>% mutate(dwc_verbatimLatitude = y) 
```
### verbatimLongitude

```{r}
event %<>% mutate(dwc_verbatimLongitude = x) 
```

### verbatimCoordinateSystem

```{r}
event %<>% mutate(dwc_verbatimCoordinateSystem = "Lambert coordinates" ) 
```

### verbatimSRS

```{r}
event %<>% mutate(dwc_verbatimSRS = "Belgian Datum 1972" ) 
```


### decimalLatitude and decimalLongitude

```{r}
event <- 
  event %>%
    st_as_sf(coords = c("x", "y"), crs = 31370) %>%
    st_transform(crs = 4326) %>%
    st_coordinates() %>%
    as.data.frame() %>%
    rename(dwc_decimalLongitude = X,
           dwc_decimalLatitude = Y) %>%
    bind_cols(event) %>%
    relocate(colnames(event))
head(event)
```

## geodeticDatum

```{r}
event %<>% mutate(dwc_geodeticDatum = "WGS84" ) 
```

## coordinateUncertainty

```{r}
occurrence %<>% mutate(dwc_coordinateUncertainty = "30") 
```


### scientificName



### kingdom

Inspect values:

```{r}
occurrence %>%
  group_by(domein) %>%
  count()
```

Map values:

```{r}
occurrence %>% mutate(dwc_kingdom = case_when(
  domein == "Dier" ~ "Animalia",
  domein == "Plant" ~ "Plantae"
))
```


### taxonRank

Inspect values:

```{r}
taxon %>%
  group_by(rankmarker) %>%
  count()
```

Map values by recoding to the [GBIF rank vocabulary](http://rs.gbif.org/vocabulary/gbif/rank_2015-04-24.xml):

```{r}
taxon %<>% mutate(dwc_taxonRank = recode(rankmarker,
  "agg."      = "speciesAggregate",
  "infrasp."  = "infraspecificname",
  "sp."       = "species",
  "var."      = "variety",
  .default    = "",
  .missing    = ""
))
```

Inspect mapped values: 

```{r}
taxon %>%
  group_by(rankmarker, dwc_taxonRank) %>%
  count()
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = "my_nomenclaturalCode") # e.g. "ICZN"
```

## Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
```

# Distribution extension

## Pre-processing

Create a dataframe with all data:

```{r}
distribution <- input_data
```

## Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id)
```

### locality

Inspect values:

```{r}
distribution %>%
  group_by(country_code, locality) %>%
  count()
```

Map values to `locality` if provided, otherwise use the country name:

```{r}
distribution %<>% mutate(dwc_locality = case_when(
  !is.na(locality) ~ locality,
  country_code == "BE" ~ "Belgium",
  country_code == "GB" ~ "United Kingdom",
  country_code == "MK" ~ "Macedonia",
  country_code == "NL" ~ "The Netherlands",
  TRUE ~ "" # In other cases leave empty
))
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(country_code, locality, dwc_locality) %>%
  count()
```

### countryCode

Inspect values:

```{r}
distribution %>%
  group_by(country_code) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_countryCode = country_code) 
```

### occurrenceStatus

Inspect values:

```{r}
distribution %>%
  group_by(occurrence_status) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceStatus = occurrence_status) 
```

### threatStatus

Inspect values:

```{r}
distribution %>%
  group_by(threat_status) %>%
  count()
```

Map values by recoding to the [IUCN threat status vocabulary](http://rs.gbif.org/vocabulary/iucn/threat_status.xml):

```{r}
distribution %<>% mutate(dwc_threatStatus = recode(threat_status,
  "endangered" = "EN",
  "vulnerable" = "VU"
))
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(threat_status, dwc_threatStatus) %>%
  count()
```

### source

Inspect values:

```{r}
distribution %>%
  group_by(source) %>%
  count() %>%
  head() # Remove to show all values
```

Map values:

```{r}
distribution %<>% mutate(dwc_source = source) 
```

### occurrenceRemarks

Inspect values:

```{r}
distribution %>%
  group_by(remarks) %>%
  count() %>%
  head() # Remove to show all values
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceRemarks = remarks) 
```

## Post-processing

Only keep the Darwin Core columns:

```{r}
distribution %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
```

Preview data:

```{r}
distribution %>% head()
```

Save to CSV:

```{r}
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
```
