---
title: "Fetch data of RATO vzw"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

Load libraries:

```{r}
library(httr)     # generic webservice package
library(ows4R)    # interface for OGC webservices
library(here)     # to work with paths
library(readr)    # to read csv files
library(safer)    # to handle encryption and decryption
library(dplyr)    # to do data wrangling
```

# Fetch data from WFS

WFS link:

```{r link}
wfs_rato <- "https://gis.oost-vlaanderen.be/server/rest/services/RATO/RATO_Public_Data/MapServer/WFSServer"
wfs_rato <- "https://gis.oost-vlaanderen.be/server/services/RATO/RATO_Public_Data/MapServer/WFSServer"
```

Create WFS client:

```{r create_client}
rato_client <- ows4R::WFSClient$new(wfs_rato, serviceVersion = "2.0.0")
```

Overview of the layers available:

```{r layers}
layers <- rato_client$getFeatureTypes(pretty = TRUE)
print(layers)
```

Request data from layer `Waarnemingen_Public:Invasieve_uitheemse_soorten`:

```{r request}
url <- httr::parse_url(wfs_rato)
url$query <- list(service = "wfs",
                  request = "GetFeature",
                  typename = "RATO_Public_Data",
                  outputFormat = "CSV"
)
request <- httr::build_url(url)
```

Download requested data as csv and save them  as `rato_data.csv`:

```{r download_data}
file_path <- here::here("data", "raw", "rato_data.csv")
httr::GET(url = request, write_disk(file_path, overwrite = TRUE))
```

## Remove column `GmlID`

The column `GMlID` is not present in the database of RATO vzw: it is added by the WFS service as a progressive number with the layer name as prefix. Removing this column from the raw data would increase the effectiveness of the git file change history tool, e.g. in case new observations are added at a certain row in the file.

```{r remove_GmlID}
# read raw occurrences
raw_occurrences <- readr::read_csv(
  file = file_path,
  na = "",
  col_types = readr::cols(.default = readr::col_character())
)

# remove GmlId column
raw_occurrences <- subset(raw_occurrences, select = -GmlId)
```
## Write raw_occurrences with Opmerkingen encrypted
### Fetch key
The key for encryption/decryption is stored as an environmental variable
```{r fetch secret}
secret_key <- Sys.getenv("opmerkingen_secret")
```
## Encrypt the Opmerkingen fields

The `OJECTID` column is appended to the `Opmerkingen` field so even empty values will result in a different encrypted string.
```{r function to encrypt column}
encrypt_column <- function(column_to_encrypt, id_col, key) {
  purrr::map2_chr(id_col, column_to_encrypt, ~ safer::encrypt_string(
    key = key,
    ascii = TRUE,
    string = paste(
      sep = "|",
      .x,
      stringr::str_replace_na(.y, replacement = "")
    )
  ))
}
```

```{r encrypt sensitive fields}
raw_occurrences <-
  raw_occurrences %>% 
  mutate(
    Opmerkingen =
      encrypt_column(Opmerkingen, OBJECTID, key = secret_key),
    Opmerkingen_admin =
      encrypt_column(Opmerkingen_admin, OBJECTID, key = secret_key)
  )
```


```{r save raw_occurrences}
# save raw occurrences (overwrite file)
readr::write_csv(
  x = raw_occurrences,
  file = file_path,
  na = "",
)
```
