---
title: "Fetch data of RATO vzw"
author:
- Damiano Oldoni
- Pieter Huybrechts
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

Load libraries:

```{r}
library(here)     # to work with paths
library(readr)    # to read csv files
```

Load helper functions: 

```{r load helper functions}
here::i_am("inst/source/fetch_data.Rmd")
devtools::load_all(path = here::here())
```
# Fetch data from the MapService
The current setup doesn't support authentication on a WFS layer. So, as a workaround, we can send a number of queries to get the data out. 

1) List all the object ids
2) Query all fields for these object ids

The query service recommends not requesting over 1000 objects in a single request. So we have to request the objects in batches and combine the responses.

## Get a list of objects to fetch

```{r}
all_object_ids <- list_object_ids(token = get_token())
message(glue::glue("Expecting {length(all_object_ids)} records"))
```

## Split the list of objects into batches
We can't request all the objects at once

```{r}
batch_size <- 100 # having some inconsistent results with higher batch sizes
batched_object_ids <- 
  split(all_object_ids, ceiling(seq_along(all_object_ids)/batch_size))
```

## Fetch the batches and combine the result
```{r Fetch the records in batches}
batched_responses <- batched_object_ids %>% 
  purrr::map(get_objects, .progress = TRUE, token = get_token())

raw_occurrences <- purrr::list_rbind(batched_responses)
```

### Check that we got the number of records we expect

```{r}
assertthat::assert_that(
  length(all_object_ids) == nrow(raw_occurrences)
)
```

## Convert date fields
Esri reports it's dates as miliseconds since 1970-01-01
```{r}
raw_occurrences <-
  raw_occurrences %>%
  dplyr::mutate(dplyr::across(
    dplyr::contains("Datum"),
    ~ as.POSIXct(.x / 1000, origin = "1970-01-01")
  ))
```

## Create object for mapping to use

```{r}
raw_data <- raw_occurrences
```
